|

### 2023-11-03 talks/midas2023/gerding - forecast evaluation via social utility.qmd
separate plotting code into talks/plots/

---
title: |
  | Infectious Disease Forecast Evaluation Via Social Utility: Allocation Scores
subtitle: MIDAS 2023
author: Aaron Gerding, Nick Reich, Ben Rogers, Evan Ray
institute: "School of Public Health and Health Sciences, UMass, Amherst"
format: 
  beamer:
    navigation: horizontal
    fontsize: 9pt
    header-includes: |
      \titlegraphic{
      \begin{minipage}{0.4\linewidth}
          \includegraphics[width=\linewidth]{reichlab.png}
      \end{minipage}%
      \hspace{0.06\linewidth}
      \begin{minipage}{0.4\linewidth}
          \includegraphics[width=\linewidth]{C19FH.png}
      \end{minipage}%
      }
  # revealjs: 
  #   width: 1050
  #   height: 700
# revealjs-plugins:
#   - attribution
editor: visual
# css: styles.css
---

## The Role of ID forecasting Hubs

Outbreak forecasting hubs developed to **"inform public health responses"**

Basic examples of **responses** are **allocation decisions** regarding

-   medical supplies
-   facility capacity
-   personnel
-   R&D funding
-   public tolerance for NPIs

## Basic Hub Products

![](plots/p_VA_CU_IHME.png)

@@

A **scoring rule** $S$ assigns a number $S(F,y)$ to the forecast $F$ of $Y$ when $Y=y$.

$S$ if **proper** if $\displaystyle{E_{F}[S(F,Y)] \leq E_{F}S(G,Y)]}$ for all $F, G$

-   ensures that honesty maximizes expected score

Hub goal of increasing social welfare by ranking and combining forecasts to best support PH responses seems to require use of proper scores.

Current standard is the **Weighted Interval Score** (a discrete version of the **Continuous Ranked Probability Score**)

-   adopted for convenient scoring of quantile forecasts.

#### A central goal in design of scoring rules for social welfare

::: box
Create scenario where the forecaster believes maximizing her own expected utility aligns with leading the DM toward actions that maximize social utility.
:::

## Bayes Acts

Scoring rules can be cast in decision-theoretic framework of **Bayes acts**.

::: box
The **Bayes act** relative to a forecast $Y \sim F$ is the action $x^F$ by which a DM will minimize **expected loss** $E_F[l(x,Y)]$.
:::

-   loss $l(x,y)$ is the **consequence** of experiencing $y$ having prepared with $x$

::: box
Key point: $S(F,y) := l(x^F, y)$ is an *automatically proper* scoring rule.
:::

#### Our basic example:

-   $x$ and $y$ are vectors in $\mathbb{N}_0^{52}$
-   $y =$ number of severe cases in US states and territories
-   $l(x,y) =$ unmet need when $x$ beds allocated and $y$ severe cases occur

@@

![](plots/iter_combo.png)

@@

![](plots/nat_hosps.png)

@@

![](plots/peak_alloscore.png)

@@

::: {.column width="60%"}
![](plots/peak_alloscore_min.png)
:::
:::

@@

![](plots/p_wis_v_alloscore.png)

@@

![](plots/mvbucky_dist_alloc.png)

>>>>

---
title: |
  | Infectious Disease Forecast Evaluation Via Social Utility: Allocation Scores
subtitle: MIDAS 2023
author: Aaron Gerding, Nick Reich, Ben Rogers, Evan Ray
institute: |
 | School of Public Health and Health Sciences, UMass, Amherst
 | Department of Biostatistics and Epidemiology
format: 
  beamer:
    keep-tex: true
    navigation: horizontal
    fontsize: 9pt
    header-includes: |
      \titlegraphic{
      \begin{minipage}{0.4\linewidth}
          \includegraphics[width=\linewidth]{reichlab.png}
      \end{minipage}%
      \hspace{0.06\linewidth}
      \begin{minipage}{0.4\linewidth}
          \includegraphics[width=\linewidth]{C19FH.png}
      \end{minipage}%
      }
  # revealjs: 
  #   width: 1050
  #   height: 700
# revealjs-plugins:
#   - attribution
editor: visual
# css: styles.css
---

## Outbreak Forecast Hubs and **Informed** Resource Allocation

Hubs developed to **"inform public health responses"**

-   such as how resources are **allocated** among locations

-   e.g., medical supplies, facility capacity, personnel

![](../plots/output/p_VA_JHU_MUNI.jpeg)

@@

Hubs strive to rank and combine forecasts so as to optimize social welfare via the public health decisions that forecasts inform

-   uncertainty quantification, and therefore, probabilistic forecasting essential

-   basic strategy: use a **scoring rule** $S$ which assigns a loss $S(F,y)$ when a probabilistic forecast $F$ of $Y$ is chosen and $Y=y$ is observed.

-   optimizing welfare requires that forecasters say what they believe; so $S$ should be **proper** meaning $\displaystyle{E_{F}[S(F,Y)] \leq E_{F}S(G,Y)]}$ for all $F, G$

Current standard is the **Weighted Interval Score** (discrete CRPS)

-   adopted largely for convenient scoring of quantile forecasts.

## Tools from decision theory

### A central goal in design of scoring rules

Tie forecast scores directly to the benefits to society of the decisions they inform

Key tools from decision theory for linking success/failure of forecast-informed policy actions to scoring rules:

-   Let $l(x,y)$ be the **loss** of experiencing $y$ after taking policy action $x$

-   The **Bayes risk** of a forecast $Y \sim F$ is $\min_x E_F[l(x,Y)]$

-   A **Bayes act** for $F$ is an action $x^F$ that attains this minimum

-   Losses from Bayes acts define an *automatically proper* scoring rule

$$
S(F,y) := l(x^F, y)
$$

**Note:** Scoring rules are really only a means to an end.  It is the expected score, in this case the Bayes risk, which characterize the value a forecast adds for a decision maker.  Scoring rule sample averages estimate the expected score.

## A new scoring rule via a new loss function for constrained actions

### Our basic example:

-   $x$ and $y$ are vectors in $\mathbb{N}_0^{52}$

-   $y =$ number of severe cases in US states and territories

-   $l(x,y) =$ unmet need when $x$ beds allocated and $y$ severe cases occur

@@

![](../plots/output/iter_combo.png)

@@

![](../plots/output/nat_hosps.png)

@@

![](../plots/output/peak_alloscore.png)

@@

::: {.column width="60%"}
![](../plots/output/peak_alloscore_min.png)
:::
:::

@@

![](../plots/output/p_wis_v_alloscore.png)

@@

![](../plots/output/mvbucky_dist_alloc.png)
